Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-06-26 Experiments.tex
down to 10!
Negative spaces fixing
refinement, reduction and reviewers..

\begin{figure}[!htb]
        \includegraphics[height=4.5cm,width=\columnwidth]{./figures/NGE/weak1.pdf}
    \caption{Weak scalability: average pilot duration, average duration of one AthenaMP execution, and pilot's overhead as a function of pilot sizes (200, 500, 1000 and 2000 nodes).}
\label{fig:weakScal1a}
\end{figure}

@@

We ran pilots with 256, 512, 1024 and 2048 worker nodes and 3\mtnote{2?} hours
walltime. Fig.~\ref{fig:weakScal2a} depicts the average pilot duration, the
average execution time of five sequential generations of AthenaMP, and the
corresponding overhead. We observe that the difference between the two durations
is more marked than in the previous experiments. Despite this, we notice that
the growth of the overhead is consistent with the increment of the number of
tasks per node for pilots with 256, 512 and 1024 worker nodes, and less than
linear for the pilot with 2048 worker nodes.

\begin{figure}[!htb]
        \includegraphics[height=4.5cm,width=\columnwidth]{./figures/NGE/weak2.pdf}
    \caption{Weak scalability with multiple generations: average pilot
    duration, average duration of sequential AthenaMP executions, and
    pilot's overhead for pilot with 256, 512, 1024 and 2048 nodes.}
\label{fig:weakScal2a}
\end{figure}

@@

The last experiments  study strong scalability by running the same number of
tasks for different pilot sizes. We used 2048 AthenaMP instances and  pilots
with 256, 512, 1024 and 2048 nodes. Thus, the number of AthenaMP generations is
equal to eight times the size of the smallest pilot and corresponds to the size
of the largest pilot. As a consequence, the number of consecutive generations of
AthenaMP decreases with the pilot size by generating different dynamics within
the pilots. These experiments are designed to investigate whether pilot overhead
is affected by the degree of concurrency within the pilot and/or the number of
tasks. Each AthenaMP instance simulates sixteen events as in the previous
experiment.

@@

\begin{figure}[!htb]
        \includegraphics[height=4.5cm,width=\columnwidth]{./figures/NGE/strong.pdf}
    \caption{Strong scalability:  average pilot duration, average duration of
    sequential AthenaMP executions, and pilot's overhead for pilots with 256, 512, 1024 and 2048 nodes.}
\label{fig:strongScala}
\end{figure}

>>>>

\begin{figure}[!t]
        \includegraphics[height=4.5cm,width=\columnwidth]{./figures/NGE/weak1.pdf}
   	\vspace{-0.3in}
    \caption{Weak scalability: average pilot duration, average duration of one AthenaMP execution, and pilot's overhead as a function of pilot sizes (200, 500, 1000 and 2000 nodes).}
\label{fig:weakScal1a}
\end{figure}

@@

We ran pilots with 256, 512, 1024 and 2048 worker nodes and 3\mtnote{2?} hours
walltime. Fig.~\ref{fig:weakScal2a} depicts the average pilot duration, the
average execution time of five sequential generations of AthenaMP, and the
corresponding overhead. The difference between the two durations
is more marked than in the previous experiments. Despite this, we notice that
the growth of the overhead is consistent with the increment of the number of
tasks per node for pilots with 256, 512 and 1024 worker nodes, and less than
linear for the pilot with 2048 worker nodes.

\begin{figure}[!t]
        \includegraphics[height=4.5cm,width=\columnwidth]{./figures/NGE/weak2.pdf}
        \vspace{-0.3in}
    \caption{Weak scalability with multiple generations: average pilot
    duration, average duration of sequential AthenaMP executions, and
    pilot's overhead for pilot with 256, 512, 1024 and 2048 nodes.}
\label{fig:weakScal2a}
\end{figure}

%\vspace{-0.2in}

@@

We investigate strong scalability by running the same number of
tasks for different pilot sizes. We used 2048 AthenaMP instances and  pilots
with 256, 512, 1024 and 2048 nodes. Thus, the number of AthenaMP generations is
equal to eight times the size of the smallest pilot and corresponds to the size
of the largest pilot. As a consequence, the number of consecutive generations of
AthenaMP decreases with the pilot size by generating different dynamics within
the pilots. These experiments are designed to investigate whether pilot overhead
is affected by the degree of concurrency within the pilot and/or the number of
tasks. Each AthenaMP instance simulates sixteen events as in the previous
experiment.

@@

\begin{figure}[!t]
        \includegraphics[height=4.5cm,width=\columnwidth]{./figures/NGE/strong.pdf}
        \vspace{-0.3in}
    \caption{Strong scalability:  average pilot duration, average duration of
    sequential AthenaMP executions, and pilot's overhead for pilots with 256, 512, 1024 and 2048 nodes.}
\label{fig:strongScala}
\end{figure}

\vspace{-0.05in}

### 2017-06-26 conclusion.tex
refinement, reduction and reviewers..

We characterized the state of practice by evaluating the efficiency, scalability
and reliability of both PanDA Broker and AthenaMP as deployed on Titan
(\S\ref{sec:panda_titan}). Our characterization highlighted the strengths
and limitations of the current design and implementation: PanDA Brokers enable
the sustained execution of millions of simulations per week but further work is
required to optimize its efficiency and reliability (\S\ref{ssec:broker_titan}).
PanDA Brokers support the concurrent execution of multiple AthenaMP instances,
enabling each AthenaMP to perform the concurrent execution of up to 16 Geant4
simulators. Nonetheless, our characterization showed how improving I/O
performance could reduce overheads (\S\ref{ssec:athenamp_titan}), increasing the
overall utilization of Titan's backfill availability.

@@

The 52M core hours used by ATLAS, via PanDA, is over 2\% of the total
utilization on Titan over the same period, bringing the time-averaged
utilization of Titan to be consistently upwards of 90\%. Given that the
average average utilization of most other leadership class machines is less
(e.g., NSF's flagship Blue Waters the average utilization fluctuates between
60-80\% (see XDMoD\cite{bw-sucks})) there is ample headroom for similar
approaches elsewhere. These unprecedented efficiency gains aside, this work is
just a starting point towards more effective operational models for future
leadership and online analytical platforms~\cite{foap-url}. These platforms
will have to support ever increasing complex workloads with varying models for
dynamic resource federation. Recent efforts with Google Cloud for the CMS
experiment -- HEPCloud~\cite{hepcloud,googlehep} represent similar yet
different approaches to dynamic resource federation. As the type,
heterogeneity and complexity of future platforms  (e.g., Exascale, Commercial
Clouds) for science increases, there will be a greater need and emphasis on
abstractions to ensure WMS can both utilize as well as support evolving
platforms. Our experience demonstrated the advantages arising from an
abstractions based approach \textendash{} in particular scalable implementations
of pilot abstraction.

>>>>

We characterized the experience by evaluating the efficiency, scalability
and reliability of both PanDA Broker and AthenaMP as deployed on Titan
(\S\ref{sec:panda_titan}). Our characterization highlighted the strengths
and limitations of the current design and implementation: PanDA Brokers enable
the sustained execution of millions of simulations per week but further work is
required to optimize its efficiency and reliability (\S\ref{ssec:broker_titan}).
PanDA Brokers support the concurrent execution of multiple AthenaMP instances,
enabling each AthenaMP to perform the concurrent execution of up to 16 Geant4
simulators. Nonetheless, our characterization showed how improving I/O
performance could reduce overheads (\S\ref{ssec:athenamp_titan}), increasing the
overall utilization of Titan's backfill availability.

@@

The 52M core hours used by ATLAS, via PanDA, is over 2\% of the total
utilization on Titan over the same period, bringing the time-averaged
utilization of Titan to be consistently upwards of 90\%. Given that the
average average utilization of most other leadership class machines is less
(e.g., NSF's flagship Blue Waters the average utilization fluctuates between
60-80\% (see XDMoD\cite{bw-sucks})) there is ample headroom for similar
approaches elsewhere. These unprecedented efficiency gains aside, this work is
just a starting point towards more effective operational models for future
leadership and online analytical platforms~\cite{foap-url}. These platforms
will have to support ever increasing complex workloads with varying models for
dynamic resource federation.

% Recent efforts with Google Cloud for the CMS
% experiment -- HEPCloud~\cite{hepcloud,googlehep} represent similar yet
% different approaches to dynamic resource federation. As the type,
% heterogeneity and complexity of future platforms  (e.g., Exascale, Commercial
% Clouds) for science increases, there will be a greater need and emphasis on
% abstractions to ensure WMS can both utilize as well as support evolving
% platforms. Our experience demonstrated the advantages arising from an
% abstractions based approach \textendash{} in particular scalable implementations
% of pilot abstraction.

### 2017-06-26 introduction.tex
refinement, reduction and reviewers..

The Large Hadron Collider (LHC) was created to explore the fundamental
properties of matter for the next decades. Multiple experiments at LHC have
collected and distributed hundreds of petabytes of data worldwide to hundreds of
computer centers. Thousands of physicists analyze petascale data volumes daily.
The detection of the Higgs Boson in 2013 speaks to the success of the detector
and experiment design, as well as the sophistication of computing systems
devised to analyze the data.

Historically, the computing systems used by LHC experiments consisted of the
federation of hundreds to thousands of distributed resources, %\textemdash{}
ranging in scale from small to mid-size resource~\cite{foster2003grid}. Although
the workloads to be executed are comprised of tasks that are independent of each
other, the management of the distribution of workloads across many heterogeneous
resources to ensure the effective utilization of resources and efficient
execution of workloads presents non-trivial challenges.

@@

As the LHC prepares for Run 3 in $~\approx$ 2022 and the high-luminosity era
(Run 4), it is anticipated that the data volumes that will need analyzing will
increase by factors of 10-100 compared to the current phase (Run 2). Data will
be larger in volume but will also require more sophisticated computational
processing. In spite of the impressive scale of the ATLAS distributed computing
system, demand for computing systems will significantly outstrip current and projected supply. %(availability).

There are multiple levels at which this problem needs to be addressed: the
utilization of emerging parallel architectures (e.g., platforms); algorithmic
and advances in analytical methods (e.g., use of Machine Learning); and the
ability to exploit different platforms (e.g., clouds and supercomputers).

This paper is a case study of how the ATLAS experiment has ``broken free'' of
the traditional computational approach of high-throughput computing on
distributed resources to embrace new platforms, in particular high-performance
computers. Specifically, we discuss the experience of integrating PanDA WMS with
a DOE leadership computing facility called Titan to reach sustained production
scales of approximately 51M core-hours a year. We also discusses the
investigation of a task execution runtime system on Titan, that is based on the
pilot-abstractions and which allows advanced execution modes of the ATLAS
workload as well as the enhanced support for heterogeneous workloads such as
molecular dynamics.

@@

This state-of-practice paper provides three main contributions:  (i) a critical
evaluation of the many design and operational considerations that have been
taken to support the sustained, scalable and production usage of Titan for
historically high-throughput workloads; (ii) a preliminary characterization of a
next generation executor (NGE) for PanDA to support non-traditional
heterogeneous workloads and execution modes;  and (iii) early lessons and
guidance as the community looks forward to designing the next generation of
online analytical platforms~\cite{foap-url}, for how current and future
experimental and observational systems can be integrated with production
supercomputers in a general and extensible manner.

>>>>

The Large Hadron Collider (LHC) was created to explore the fundamental
properties of matter. Multiple experiments at LHC have
collected and distributed hundreds of petabytes of data worldwide to hundreds of
computer centers. Thousands of physicists analyze petascale data volumes daily.
The detection of the Higgs Boson in 2013 speaks to the success of the detector
and experiment design, as well as the sophistication of computing systems
devised to analyze the data.

Historically, the computing systems used by LHC experiments consisted of the
federation of hundreds to thousands of distributed resources, %\textemdash{}
ranging in scale from small to mid-size resource~\cite{foster2003grid}. Although
the workloads to be executed are comprised of tasks that are independent of each
other, the management of the distribution of workloads across many heterogeneous
resources, the effective utilization of resources and efficient
execution of workloads presents non-trivial challenges.

@@

As the LHC prepares for Run 3 in $~\approx$ 2022 and the high-luminosity era
(Run 4), it is anticipated that the data volumes that will need analyzing will
increase by factors of 10-100 compared to the current phase (Run 2). Data will
be larger in volume but will also require more sophisticated computational
processing. In spite of the impressive scale of the ATLAS distributed
computing system, demand for computing systems will significantly outstrip
current and projected supply.  There are multiple levels at which this problem
needs to be addressed: the utilization of emerging parallel architectures
(e.g., platforms); algorithmic and advances in analytical methods (e.g., use
of Machine Learning); and the ability to exploit different platforms (e.g.,
clouds and supercomputers).

This paper represents the experience of how the ATLAS experiment has ``broken
free'' of the traditional computational approach of high-throughput computing
on distributed resources to embrace new platforms, in particular  high-
performance computers. Specifically, we discuss the experience of integrating
PanDA WMS with a US DOE leadership computing facility called Titan to reach
sustained production scales of approximately 51M core-hours a year. We also
discusses the investigation of a pilot-abstraction based task execution
runtime system on Titan, which allows advanced execution modes of the ATLAS
workload as well as the enhanced support for heterogeneous workloads, 
e.g., molecular dynamics.

@@

This experience paper provides three main contributions:  (i) a critical
evaluation of the many design and operational considerations that have been
taken to support the sustained, scalable and production usage of Titan for
historically high-throughput workloads; (ii) a preliminary characterization of a
next generation executor (NGE) for PanDA to support non-traditional
heterogeneous workloads and execution modes;  and (iii) early lessons and
guidance as the community looks forward to designing the next generation of
online analytical platforms~\cite{foap-url}, for how current and future
experimental and observational systems can be integrated with production
supercomputers in a general and extensible manner.

### 2017-06-26 paper.tex
refinement, reduction and reviewers..

>>>>

\author{
    \IEEEauthorblockN{
        Alessio Angius\IEEEauthorrefmark{1}\IEEEauthorrefmark{2},
        Danila Oleynik\IEEEauthorrefmark{1}\IEEEauthorrefmark{5},
        Sergey Panitkin\IEEEauthorrefmark{1}\IEEEauthorrefmark{4},
        Matteo Turilli\IEEEauthorrefmark{1}\IEEEauthorrefmark{2},\\
        Kaushik De\IEEEauthorrefmark{5},
        Alexei Klimentov\IEEEauthorrefmark{4},
        Sarp H. Oral\IEEEauthorrefmark{6},
        Jack C. Wells\IEEEauthorrefmark{6} and
        Shantenu Jha\IEEEauthorrefmark{2}
    }
    \IEEEauthorblockA{
        \IEEEauthorrefmark{1}First authors, alphabetical order
    }
    \IEEEauthorblockA{
        \IEEEauthorrefmark{2}Department of Electric and Computer Engineering, Rutgers University, NJ, USA
    }
    \IEEEauthorblockA{
        \IEEEauthorrefmark{4}Physics Department, Brookhaven National Laboratory, NY, USA
    }
    \IEEEauthorblockA{
        \IEEEauthorrefmark{5}Physics Department, University of Texas Arlington, TX, USA
    }
    \IEEEauthorblockA{
        \IEEEauthorrefmark{6}Oak Ridge Leadership Computing Facility, Oak Ridge National Laboratory, TN, USA
    }
}
